---
title: "pogos -- PharmacOGenomics Ontology Support"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{pogos -- simple interface to bhklab PharmacoDB with emphasis on ontology}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

bhklab's PharmacoDB unites multiple drug sensitivity databases.
A REST API is available and interfaces to some basic endpoints are
provided in this package.  We also provide some basic
support for working with the Cell Line Ontology and ChEBI.

```{r setup,echo=FALSE}
suppressMessages({
suppressPackageStartupMessages({
library(pogos)
library(S4Vectors)
library(DT)
})
})
```

# Some stored reference data

We precomputed tables for cell lines, compounds, and tissues.
These are available as DataFrame instances as
`cell_lines_v1`, `compounds_v1`, and `tissues_v1`, in
the data element of the package.

Numerical ids are defined by the PharmacoDB.

## Cell lines
```{r lkk, echo=FALSE}
data(cell_lines_v1)
datatable(as.data.frame(cell_lines_v1), options=list(lengthMenu=c(3,5,10,50,100)))
```

## Compounds
```{r lkccc, echo=FALSE}
data(compounds_v1)
datatable(as.data.frame(compounds_v1), 
   options=list(lengthMenu=c(3,5,10,50,100)))
```

# Bridging to Cell Line Ontology

The following identifiers are available in the PharmacoDB cell lines:
```
c("143B", "1321N1", "184B5")
```
What are they?  What organs do they represent?  One way
to proceed is through parent terms in Cell Line Ontology.
```{r lkoo}
library(ontoProc)
clo = getCellLineOnto()
minds = which(clo$name %in% c("143B cell", "1321N1 cell", "184B5 cell"))
tags = clo$id[minds]
clo$name[ unlist(clo$parent[tags]) ]
```
Now: how can we connect the anatomic vocabulary in these terms
to formal anatomical terms in UBERON or CARO?

# Bridging to ChEBI

This will involve more work.  We have the compounds in PharmacoDB.
How many match the compound names in Chebi? 
```{r lkcom}
chl = getChebiOnto()
allch = chl$name
mm = allch[match(compounds_v1[,2], allch, nomatch=0)]
round(length(mm)/nrow(compounds_v1),2) # not high
datatable(data.frame(id=names(mm), comp=as.character(mm)))
```

Direct matches are relatively rare.  We also search the
synonym data in ChEBI

```{r srchsym,cache=TRUE}
notin = setdiff(compounds_v1$name, mm)
#library(BiocParallel)
#register(MulticoreParam())
library(parallel)
options(mc.cores=2)
lk50 = mclapply(notin[1:50], function(x) fastGrep(x, chl, "synonym")[1])
uu = unlist(lk50)
mean(is.na(uu)) # recovery not impressive
names(uu) = notin[1:50]
inds = na.omit(uu)
```
```{r lktabb}
datatable(data.frame(id=chl$id[inds], given=names(inds), chebi=chl$name[inds]))
```

