---
title: "pogos -- PharmacOGenomics Ontology Support"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{pogos -- simple interface to bhklab PharmacoDB with emphasis on ontology}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc_float: yes
---

# Introduction

bhklab's PharmacoDB unites multiple drug sensitivity databases.
A REST API is available and interfaces to some basic endpoints are
provided in this package.  We also provide some basic
support for working with the Cell Line Ontology and ChEBI.

We note that detailed RESTful interrogation of many ontologies
is supported by the `r Biocpkg("rols")` package.  In the long
run we would like to make use of the EBI Ontology Lookup System
as a central resource of this package, but to develop concepts
at these early stages, static images of ontologies are employed.

```{r setup,echo=FALSE}
suppressMessages({
suppressPackageStartupMessages({
library(pogos)
library(S4Vectors)
library(DT)
})
})
```

## Coverage of pharmacogenomic compound names by ChEBI

As an illustration of long-term intention of this package, consider
the following snapshot of a slice of ChEBI.

```{r doch,fig=TRUE}
library(ontoProc)
cc = getChebiOnto()
library(ontologyPlot)
onto_plot(cc, 
  c("CHEBI:134109", "CHEBI:61115", "CHEBI:75603", "CHEBI:37699", 
       "CHEBI:68481", "CHEBI:71178"), fontsize=24, shape="rect")
```
It is natural to map the compounds studied in pharmacogenomics to these
compound classes as a step towards understanding common and distinct
mechanisms of action.  But in the following display, we see
that some common chemotherapeutic agents are characterized in very
generic terms:

```{r lkmo, fig=TRUE}
onto_plot(cc, c("CHEBI:134109", "CHEBI:61115", "CHEBI:75603", "CHEBI:37699", 
   "CHEBI:68481", "CHEBI:71178", "CHEBI:52172", "CHEBI:64310", 
   "CHEBI:63632", "CHEBI:45863"), fontsize=28, shape="rect")
```

Should paclitaxel be identified as an "organic molecular entity"
with no additional detail?  Is there additional information in
ChEBI or related ontologies to help organize information on
related compounds?

## Relationships among cell lines

Cell line nomenclature is often unnatural.  The Cell Line
Ontology should be helpful in systematizing cell lines
used in pharmacogenomics.  Nevertheless there is
some apparent lack of consistency.

```{r ana,fig=TRUE}
clo = getCellLineOnto()
smoc = c("CLO:0000517", "CLO:0000560", "CLO:0000563", "CLO:0001072", 
"CLO:0001088", "CLO:0001138", "CLO:0007606")
onto_plot(clo, smoc)
```

The MCF7 line should probably be annotated as related to human
mammary tissue.

# Some stored reference data

We precomputed tables for cell lines, compounds, and tissues.
These are available as DataFrame instances as
`cell_lines_v1`, `compounds_v1`, `datasets_v1`, and `tissues_v1`, in
the data element of the package.

Numerical ids are defined by the PharmacoDB.

## Cell lines
```{r lkk, echo=FALSE}
data(cell_lines_v1)
datatable(as.data.frame(cell_lines_v1), options=list(lengthMenu=c(3,5,10,50,100)))
```

## Compounds
```{r lkccc, echo=FALSE}
data(compounds_v1)
datatable(as.data.frame(compounds_v1), 
   options=list(lengthMenu=c(3,5,10,50,100)))
```

## Datasets
```{r lkccc2, echo=FALSE}
data(datasets_v1)
datatable(as.data.frame(datasets_v1))
```

# Working with specific elements

We want to maximize utility of institutionally curated
terminology for components of pharmacodb.

## Cell line identifiers

### Bridging to Cell Line Ontology

The following identifiers are available in the PharmacoDB cell lines:
```
c("143B", "1321N1", "184B5")
```
What are they?  What organs do they represent?  One way
to proceed is through parent terms in Cell Line Ontology.
```{r lkoo}
library(ontoProc)
clo = getCellLineOnto()
minds = which(clo$name %in% c("143B cell", "1321N1 cell", "184B5 cell"))
tags = clo$id[minds]
clo$name[ unlist(clo$parent[tags]) ]
```
TODO: how can we connect the anatomic vocabulary in these terms
to formal anatomical terms in UBERON or CARO?

### Vocabularies of anatomy

The Experimental Factor Ontology unifies various ontologies and
includes terms related to anatomic systems.  At least with the
infrastructure available in `r CRANpkg("ontologyIndex")` and
`r CRANpkg("ontologyPlot")`, it seems there are opportunities to
add meaningful connections.

```{r lkefo, fig=TRUE}
ee = getEFOOnto()
onto_plot(ee, c("UBERON:0000474", "UBERON:0000079", 
  "UBERON:0000990", "UBERON:0000995", "UBERON:0003982", 
  "UBERON:0000310", "UBERON:0002107", "UBERON:0001264", 
  "UBERON:0002113", "UBERON:0001007", "UBERON:0002530", 
  "UBERON:6007435", "UBERON:0004122", "UBERON:0001008"))
```

It is likely important to remember the distinction between an
ontology, a structured collection of concept terms, and a
curated vocabulary or ontology annotation, which covers
concept "instances".

## Compounds

### Bridging to ChEBI

This will involve more work.  We have the compounds in PharmacoDB.
How many match the compound names in Chebi? 
```{r lkcom}
chl = getChebiOnto()
allch = chl$name
mm = allch[match(tolower(compounds_v1[,2]), tolower(allch), nomatch=0)]
round(length(mm)/nrow(compounds_v1),2) # not high
datatable(data.frame(id=names(mm), comp=as.character(mm)))
```

Direct matches are relatively rare.  Attempts to use approximate
matching with agrep have so far been unproductive.  As an
example, which we do not run:

```{r srchsym, eval=FALSE}
notin = setdiff(compounds_v1$name, mm)
library(parallel)
options(mc.cores=2)
allch = tolower(allch)
notin = tolower(notin)
lk50 = mclapply(notin[1:50], function(x) { cat(x);
       order(adist(x, allch))[1:5]})
names(lk50) = notin[1:50]
aa = do.call(cbind, lk50[1:50])
ttt = t(apply(aa,2,function(x) allch[x])[1:3,])
```

# Retrieving dose-response data for cell-line/compound intersections

Given a cell-line code and a database, we can retrieve all the experiments
associated with it in that database.

CCLE for MCF-7:
```{r lkcomb}
library(rjson)
library(httr)
xx = GET("https://pharmacodb.pmgenomics.ca/api/v1/intersections/2/895/1?indent=true")
ans = fromJSON(readBin(xx$content, what="character"))
doses1 = sapply(ans[[1]]$dose_responses, function(x) x[[1]])
resp1 =  sapply(ans[[1]]$dose_responses, function(x) x[[2]])
```

The `compoundsByCell` function employs this pattern to establish
lists of experiments per cell type.
